\Large{Adversarial Training against Systematic Uncertainty}

### 2016-09-11 tex/paper.tex
save

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. In
particular, we pit $f$ against an adversary model $r := p_{\theta_r}(z |
f(X;\theta_f))$ of parameters $\theta_r$ and associated loss ${\cal
L}_r(\theta_f, \theta_r)$. This model takes  as input realizations of $f(X;
\theta_f)$, for the current value $\theta_f$ of $f$ parameters, and produces as
output probability estimates $p_{\theta_r}(z | f(X;\theta_f))$ that $f(X;
\theta_f)$ is generated from the nuisance value $z$. Intuitively, if $p(f(X;
\theta_f)|z)$ varies with $z$, then the corresponding correlation can be
captured by $r$. By contrast, if $p(f(X; \theta_f)|z)$ is invariant with $z$, as
we require, then $r$ should perform poorly and be close to random guessing.
Training $f$ such that it additionally minimizes the performance of $r$
therefore acts as a regularization towards Eqn.~\ref{eqn:criterion}.

@@

Proposition~\ref{prop:2} suggests that if at each step of
Algorithm~\ref{alg:adversarial-training} the adversary $r$ is allowed to reach
its optimum given $f$ (e.g., by setting $K$ sufficiently high) and if $f$ is
updated to improve ${\cal L}_f(\theta_f) - H(p_{Z|f(X;\theta_f)})$, then $f$
should converge to a classifier which is both optimal and pivotal, provided such
a classifier exists. On many problems of interest though, such a classifier may
not exist because the nuisance parameter directly shapes the decision boundary,
in which cases the lower bound $H(p_{Y|X}) - H(p_Z) < {\cal L}_f(\theta_f) -
H(p_{Z|f(X;\theta_f)})$ is strict: $f$ can either be an optimal classifier or a
pivotal quantity, but not both simultaneously. In this situation, it is natural
to rewrite the value function $E$  as
\begin{equation}
    E_\lambda(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r),
\end{equation}
where $\lambda \geq 0$ is a hyper-parameter controlling the trade-off between
the performance of $f$ and its independence with respect to the nuisance
parameter. Setting $\lambda$ to a large value will preferrably enforces $f$ to
be pivotal while setting $\lambda$ close to $0$ will rather constraint $f$ to be
optimal.

@@

\glnote{Define the architecture of $r$ when $Z$ is categorical ($r$ is a
standard classifier) or continuous ($r$ is e.g a 2-output NN such that the
distribution of $Z|f(X)$ is modeled by a truncated gaussian of known support). Cite Nix.}

@@

To account for systematic uncertainties, experimentalists in high energy physics
typically take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty
\glnote{add ref} by estimating $p(f(x)|z)$ with a parameterized calibration
procedure. Clearly, this classifier is however not optimal for $z \neq z_0$. In
this setting, parameterized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  As argued in
\citep{Neal:2007zz}, such classifiers can however not be used on real data since
the correct value $z$ of the nuisance often remains unknown. This is typically
not an issue in the context of parameter
inference~\citep{cranmer2015approximating}, where nuisance parameters are
marginalized out, but otherwise often limits the range of their applications. In
practice, parameterized classifiers  are also often computationally expensive to
build and evaluate. In particular, calibrating their decision fucntion, i.e.
approximating $p(f(x,z)|z)$ as a continuous function of $z$, remains an open
challenge. By contrast, constraining $f$ to be pivotal yields a classifier which
may not be optimal with respect to $Y|X,Z$, as discussed in Section~\ref{sec:theory}, but that can be
used in a wider range of applications, since knowing the correct value $z$ of
the nuisance is not necessary, and for which calibration needs to be carried out
only once, since  the dependence on the nuisance is now built in the classifier.
\glnote{Shall we also discuss the relation with \citep{Neal:2007zz} where point
estimates of the nuisance are used as inputs to $f$?}

>>>>

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. In
particular, we pit $f$ against an adversary model $r := p_{\theta_r}(z |
f(X;\theta_f))$ of parameters $\theta_r$ and associated loss ${\cal
L}_r(\theta_f, \theta_r)$. This model takes  as input realizations of $f(X;
\theta_f)$, for the current value $\theta_f$ of $f$ parameters, and produces as
output a function $p_{\theta_r}(z | f(X;\theta_f))$ modeling the conditional probability density that $f(X;
\theta_f)$ is generated from the nuisance value $z$.
\glnote{Give examples of $r$ for categorical and continuous nuisances.}
Intuitively, if $p(f(X;
\theta_f)|z)$ varies with $z$, then the corresponding correlation can be
captured by $r$. By contrast, if $p(f(X; \theta_f)|z)$ is invariant with $z$, as
we require, then $r$ should perform poorly and be close to random guessing.
Training $f$ such that it additionally minimizes the performance of $r$
therefore acts as a regularization towards Eqn.~\ref{eqn:criterion}.

@@

Proposition~\ref{prop:2} suggests that if at each step of
Algorithm~\ref{alg:adversarial-training} the adversary $r$ is allowed to reach
its optimum given $f$ (e.g., by setting $K$ sufficiently high) and if $f$ is
updated to improve ${\cal L}_f(\theta_f) - H(p_{Z|f(X;\theta_f)})$, then $f$
should converge to a classifier which is both optimal and pivotal, provided such
a classifier exists. On many problems of interest though, such a classifier may
not exist because the nuisance parameter directly shapes the decision boundary,
in which cases the lower bound $H(p_{Y|X}) - H(p_Z) < {\cal L}_f(\theta_f) -
H(p_{Z|f(X;\theta_f)})$ is strict: $f$ can either be an optimal classifier or a
pivotal quantity, but not both simultaneously. In this situation, it is natural
to rewrite the value function $E$  as
\begin{equation}
    E_\lambda(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r),
\end{equation}
where $\lambda \geq 0$ is a hyper-parameter controlling the trade-off between
the performance of $f$ and its independence with respect to the nuisance
parameter. Setting $\lambda$ to a large value will preferably enforces $f$ to
be pivotal while setting $\lambda$ close to $0$ will rather constraint $f$ to be
optimal.

@@

\glnote{Define the architecture of $r$ when $Z$ is categorical ($r$ is a
standard classifier) or continuous ($r$ is e.g a 2-output NN such that the
distribution of $Z|f(X)$ is modeled by a truncated Gaussian of known support). Cite Nix.}

@@

To account for systematic uncertainties, experimentalists in high energy physics
typically take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty
\glnote{add ref} by estimating $p(f(x)|z)$ with a parameterized calibration
procedure. Clearly, this classifier is however not optimal for $z \neq z_0$. In
this setting, parameterized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  As argued in
\citep{Neal:2007zz}, such classifiers can however not be used on real data since
the correct value $z$ of the nuisance often remains unknown. This is typically
not an issue in the context of parameter
inference~\citep{cranmer2015approximating}, where nuisance parameters are
marginalized out, but otherwise often limits the range of their applications. In
practice, parameterized classifiers  are also often computationally expensive to
build and evaluate. In particular, calibrating their decision function, i.e.
approximating $p(f(x,z)|z)$ as a continuous function of $z$, remains an open
challenge. By contrast, constraining $f$ to be pivotal yields a classifier which
may not be optimal with respect to $Y|X,Z$, as discussed in Section~\ref{sec:theory}, but that can be
used in a wider range of applications, since knowing the correct value $z$ of
the nuisance is not necessary, and for which calibration needs to be carried out
only once, since  the dependence on the nuisance is now built in the classifier.
\glnote{Shall we also discuss the relation with \citep{Neal:2007zz} where point
estimates of the nuisance are used as inputs to $f$?}
